---
title: 'Stage Two Task: Part 2'
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<style>
html { padding: 10px; } 
body { font-family: "Helvetica Neue",Helvetica,Arial,sans-serif ; font-size: 12pt ; line-height: 1.6 ; max-width: 100%; }
p { text-align: justify; }
</style>

## Part 2: Breast Cancer Data Exploration
```{r Load library and data file(s), message=FALSE, warning=FALSE}
#install.packages('reshape') to install the 'reshape' package if not yet installed
library(ggplot2)
library(dplyr)
library(reshape)

data_url <- 'https://raw.githubusercontent.com/HackBio-Internship/2025_project_collection/refs/heads/main/Python/Dataset/data-3.csv'

breastCancer_data <- read.delim(data_url, header = T, sep = ",")
```

### **c. Scatter Plot (radius vs texture)**

Goal:

-  Plot 'texture_mean' vs 'radius_mean' and color points by 'diagnosis' (M = malignant, B = benign).
```{r Scatter Plot, message=FALSE, warning=FALSE}
plot(data = breastCancer_data, 
     texture_mean~radius_mean,
     xlab = 'Radius Mean',
     ylab = 'Texture Mean',
     las = 1,
     xlim = c(0,30),
     ylim = c(0,50),
     col = c("B" = "orange2", "M" = "#3685d3")[as.factor(breastCancer_data$diagnosis)],
     pch = 19,
     cex = 0.5
     )

legend(25,50,
  legend = c("Benign", "Malignant"),
  col = c("Benign" = "orange2", "Malignant" = "#3685d3"),
  title = "Diagnosis",
  title.font = 2,# make the legend title bold
  bty = "o", #parameter to draw border: "o" fr draw box (default); "n" for no box
  pch = 15,
  cex = 0.8,
  adj = 0, #Justifies text relative to symbol	0 (left), 1 (right)
  xjust = -0.2, # Justifies box relative to X-coord	0.5 (centered)
  x.intersp = 0.8,# Space between symbol and label. Default is 1
  y.intersp = 1.2, # y.intersp	Vertical	Distance between rows of items.
  title.adj = 0.3, #Justifies the legend title	0 to 1.0 (left), 1 (right)
  box.lty = 1,
  box.lwd = 1,
  box.col = "darkgray"
  )
```

### **d. Correlation Heatmap**

Goal:

- Compute the correlation matrix of six key features:
    radius_mean, texture_mean, perimeter_mean, area_mean, smoothness_mean and compactness_mean
- Plot as a heatmap with correlation annotated.

```{r Heatmap of correlation matrix for six key features , message=FALSE, warning=FALSE}
# Create a dataframe of correlations
cor_matrix <- cor(breastCancer_data[3:8])
#print(cor_matrix) or alternatively click on the file to view it in the source panel. 

melted_cor <- melt(cor_matrix, varnames = c("var1","var2" ))

ggplot(melted_cor, aes(x = var1, y = var2, fill = value))+
  geom_tile(color = 'black')+
  geom_text(aes(label = sprintf("%.1f", value)), size = 4)+
  scale_y_discrete(limits = rev(rownames(cor_matrix)))+
  scale_fill_gradient(low = "#e7f5ff", high = "royalblue")+
  labs(
    x = '',
    y = '',
    fill = "Correlation",
    title = "Correlation Matrix of Six Key Features"
  )+
  theme_minimal()+
  theme(
     axis.text.x = element_text(angle = 90, vjust = 0.3, hjust = 1),
     axis.text = element_text(color = "black", size = 10),
     legend.title = element_text(face = "bold"),
    legend.key.height = unit(1, "null"))+
coord_fixed()
```

### **e. Scatter Plot (compactness vs smoothness)**

Goal:

-  Plot 'compactness_mean' vs 'smoothness_mean' and color points by 'diagnosis' (M = malignant, B = benign).
- Include gridlines and clear axis labels.
```{r Scatter Plot: compactness vs smoothness, message=FALSE, warning=FALSE}
ggplot(breastCancer_data, aes(x = smoothness_mean, y = compactness_mean, color = as.factor(diagnosis)))+
  geom_point(size = 1, alpha = 0.7)+
  scale_color_manual(
    labels = c("B" = "Benign","M" = "Malignant"), values = c("B" = "orange2", "M" = "#3685d3")
  )+
  coord_cartesian(xlim = c(0.05,0.2), ylim = c(0,0.35))+
  labs(
    x = "smoothness_mean",
    y = "compactness_mean",
    color = "Diagnosis",
    title = "Scatter Plot for compactness and smoothness means"
  )+
  theme_minimal()+
  theme(
    legend.background = element_rect(color = "darkgrey"),
    legend.justification = c(0,0.9),
    legend.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
    plot.title = element_text(margin = margin(t = 12, b = 12)))
```



### **f. Density Plot (area distribution)**

Goal:

-  Plot kernel density estimates (KDE) of 'area mean' for both M and B diagnosis on the same axis
- Add legend and labeled axis

```{r Density plot: kernel density estimates of area mean for both diagnosis, message=FALSE, warning=FALSE}
ggplot(breastCancer_data, aes(x = area_mean, fill = factor(diagnosis), color = factor(diagnosis)))+
  geom_density(alpha = 0.5)+
  scale_fill_manual(
    labels = c("B" = "Benign", "M" = "Malignant"), values = c("B" = "orange2", "M" = "#3685d3")
  )+
  scale_color_manual(
    labels = c("B" = "Benign", "M" = "Malignant"), values = c("B" = "orange2", "M" = "#3685d3")
  )+
  labs(
    y = "Density",
    fill = "Diagnosis",
    color = "Diagnosis",
    title = "Kernel Density Estimates of Area Mean"
    )+
  theme_minimal()+
  theme(
    panel.grid = element_blank(), 
    legend.title = element_text(face = "bold", hjust = -0.2, margin = margin(b = 6)),
    legend.background = element_rect(color = "darkgrey"),
    legend.margin = margin(5,8,8,12),
    legend.position = c(0.8, 0.9),
    legend.key = element_rect(color = NA),
    legend.key.width = unit(1, "cm"),
    legend.key.height = unit(0.3, "cm"),
    legend.key.spacing.y = unit(0.1, "cm")
  )
```

```{r}
bubble_plot_df <- read_excel("hb_stage_2.xlsx", sheet = "e")

ggplot(bubble_plot_df, aes(x = half_life, y = alpha))+
  geom_point(aes(color = stage, size = count), alpha = 0.8)+
  coord_cartesian(
    xlim = c(0,50), ylim = c(0,2)
    )+
  labs(
    x = "Half Life",
    y = "Alpha Life",
    col = "",
    size = "",
    title = "Bobble Plot of Kinetic Regimes: Alpha Life VS Half Life"
  )+
  theme_minimal()+
  theme(
    plot.margin = margin(2,20,2,20),
    plot.title = element_text(hjust = 0.5, margin = margin(t = 12, b = 12)),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    legend.position = c(1,1),
    legend.justification = c(1,1)
  )
```

