---
title: 'Stage Two Task: Part 3'
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<style>
html { padding: 10px; } 
body { font-family: "Helvetica Neue",Helvetica,Arial,sans-serif ; font-size: 12pt ; line-height: 1.6 ; max-width: 100%; }
p { text-align: justify; }
</style>

### Task 0. Orientation and data hygiene (mandatory, graded)
- Download the source paper and identify Figure 2 panels (a–g).
- Inspect the Excel file and map each sheet to a figure panel.

  Panel a: **sheet a** <br> 
  Panel b: **sheet b** <br>
  Panel c: **sheet c** <br>
  Panel d: **sheet d_1** <br>
  Panel e: **sheet e** <br>
  Panel f: **sheet f** <br>
  Panel g: **sheet g** 
  
  **NOTE:** the **&lt;br&gt;** tag is an HTML tag, it is for styling and does not special function in R!

- Install the following packages:
  - readxl
  - ggplot2
  - pheatmap
  - igraph

```{r Package installation and color function, message=FALSE, warning=FALSE}
#install.packages("readxl")
#install.packages("igraph")

#load libraries
library("readxl")
library("ggplot2")
library("pheatmap")
library("igraph")
library("dplyr")
library("stringr")

#use this function to create transparent colors
transparent_color <- function(color, percent = 50, name = NULL) {
  #      color = color name
  #    percent = % transparency
  #       name = an optional name for the color
  
  ## Get RGB values for named color
  rgb.val <- col2rgb(color)
  
  ## Make new color using input color as base and alpha set by transparency
  t.col <- rgb(rgb.val[1], rgb.val[2], rgb.val[3],
               max = 255,
               alpha = (100 - percent) * 255 / 100,
               names = name)
  
  ## Save the color
  invisible(t.col)
}

hb_pal <- c("#4e79a7", 
            "#8cd17d", 
            "#e15759", 
            "#fabfd2", 
            "#a0cbe8", 
            "#59a14f", 
            "#b07aa1", 
            "#ff9d9a", 
            "#f28e2b", 
            "#f1ce63",
            "#79706e",
            "#d4a6c8",
            "#e9e9e9",
            "#ffbe7d",
            "#bab0ac",
            "#9d7660",
            "#d37295",
            "#86bcb6",
            "#362a39",
            "#cd9942")

#test the color pallete
plot(1:length(hb_pal), 1:length(hb_pal), col = hb_pal, pch = 19)
```

### Task 1. Reproduce panel 2a: Cell-type ratio distributions
**Goal: Compare distributions across immune cell types.**

Requirements:

    1. Read sheet a
    2. Produce a boxplot of new_ratio grouped by cell_type
    3. Match:
        Orientation (rotated labels)
        Relative scaling
        Outlier visibility
        
```{r Boxplot for comparison of distributions across immune cell types , message=FALSE, warning=FALSE}
boxplot_df <- read_xlsx('hb_stage_2.xlsx', sheet = 'a')

plot1 <- ggplot(boxplot_df, aes(x = cell_type, y = new_ratio, fill = cell_type)) +
  geom_boxplot(
    #outlier parameters for outlier visibility
    notch = F,
    outlier.shape = 1, #hollow circle
    outlier.size = 0.8, #size of outliers
    outlier.stroke = 1, #thickness of hollow
    #outlier.alpha = 1, #transparency
    outlier.colour = "black", #color (or map to category)
    # linetype = "solid",
    whisker.linetype = "dashed", # makes whiskers dashed
    staplewidth = 0.3 # adds caps to whiskers
    )+
  scale_y_continuous(
       breaks = seq(0,0.5, by = 0.1),
       limits = c(0,0.5),
       expand = c(0,0)
  )+
  # this layer for relative scaling
  scale_fill_manual(values = hb_pal)+
  labs(
    x = "",
    y = "Ratio",
    title = "Cell-type Ratio Distributions"
  )+
  theme_minimal()+
  theme(
    plot.margin = margin(2,20,2,20),
    plot.title = element_text(hjust = 0.5, size = 18, margin = margin(b = 12, t = 12)),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black", size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), # this line of code for rotated labels
    axis.ticks = element_line(color = "black"),
    legend.position = "none"
  )
print(plot1)
```



### Task 2. Reproduce panel 2b: Half-life vs alpha-life scatter
**Goal: Identify kinetic regimes.**

Requirements:

    Read sheet b
    Plot log2(half_life) vs log2(alpha)
    Add:
        Vertical and horizontal cutoffs
        Color-coded subsets based on thresholds
        Labeled exemplar genes (Camp, Ccr2)

Conceptual checks:

    Why log2?
    What do the four quadrants mean?
    
    Response: log2 because it makes data symmetric and easier to interpret.
    
    Response: The four quadrants essentially group the expressions into 4 groups : expression strength across the x-axis and fold change (alpah life) along the y-axis.
    

```{r Kineetic regimes indicated with a plot of log2(half-life) VS log2(alpha), message=FALSE, warning=FALSE}
kinetics_df <- read_excel("hb_stage_2.xlsx", col_names = TRUE, sheet = "b")
### NOTE: The order of the code is important. At some point, I put this code after the code used to create the cutoff and it was not producing the correct output. Therefore, be careful when next I have to make a plot!
#Create the log2 data for both alpha and half_life
kinetics_df$log2_alpha <- log2(kinetics_df$alpha)
kinetics_df$log2_half_life <- log2(kinetics_df$half_life)

# Create vertical and horizontal cutoffs
# The code below uses base R function quantile() to group the data. The cutoff is set to a threshold of 80% for both axes
cutoff_x <- quantile(kinetics_df$log2_half_life, 0.95)
cutoff_y <- quantile(kinetics_df$log2_alpha, 0.9)

# Add the grouping for each observation in the kinetics dataframe based on the cutoff for subsequent color in the plot
kinetics_df <- kinetics_df %>%
  mutate(color_group = 
    case_when(
      log2_half_life <= cutoff_x & log2_alpha >= cutoff_y ~ "TopLeft",
      log2_half_life >= cutoff_y & log2_alpha >= cutoff_y ~ "TopRight",
      log2_half_life >= cutoff_x & log2_alpha <= cutoff_y ~ "Right",
      TRUE ~ "Other"
    )
  )

plot2 <- ggplot(kinetics_df, aes(x = log2_half_life, y = log2_alpha))+
  geom_point(aes(color = color_group), size = 1, alpha = 0.7)+
  
  # Quadrant lines
  geom_vline(xintercept = cutoff_x, linetype = "dashed")+
  geom_hline(yintercept = cutoff_y, linetype = "dashed")+
  geom_text(
    data = subset(kinetics_df, cell %in% c("Camp", "Ccr2")), 
    aes(label = cell),
    vjust = 1.5,
    hjust = 0.5,
    size = 5,
    color = "black",
    fontface = "bold"
  )+
  scale_color_manual(
    values = c(
      "TopLeft" = "#59a14f",
      "TopRight" = "#e15759",
      "Right" = "#4e79a7",
      "Other" = "#79706e"
    )
  )+
  labs(
    x = "log2(Half Life)",
    y = "log2(Alpha Life)",
    title = "Kinetic Regimes of log2(Half Life) VS log2(Alpha Life)"
  )+
  scale_y_continuous(
       breaks = seq(-10,0, by = 2),
       limits = c(-10.5,0.5),
       expand = c(0,0)
     )+
  scale_x_continuous(
       breaks = seq(-1,5, by = 1),
       limits = c(-1.5,5.5),
       expand = c(0,0)
     )+
  theme_minimal()+
  theme(
    plot.margin = margin(2,20,2,20),
    plot.title = element_text(hjust = 0.5, margin = margin(t = 12, b = 12)),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black", size = 12),
    axis.ticks = element_line(color = "black"),
    axis.text.y = element_text(angle = 90, hjust = 0.8),
    axis.title = element_text(size = 12),
    legend.position = "none"
  )
print(plot2)
```



### Task 3. Reproduce panel 2c: Heatmap across cell types and time
**Goal: Visualize temporal structure across immune compartments.**

Requirements:

    Read sheet c
    Convert to matrix correctly
    Build column annotations for:
        CellType
        Time
    Cluster rows only, not columns

Key thinking check:

    Why cluster genes but not time?
    
    Response: genes are clustered because the intent is to check the expression of genes in the various cells over time. If time is clustered, it will combine the expression of th genes which will mask the expression of each gene which is the original intent of the task.

```{r Heatmap across cell types and time, message=FALSE, warning=FALSE}
heatmap_df <- read_excel("hb_stage_2.xlsx", sheet = "c")

#Create column annotation
col_annotation_b <- data.frame(
  CellType = str_extract(colnames(heatmap_df[-1]), "^[A-Za-z]+(?=n\\d)"),
  Time = str_extract(colnames(heatmap_df[-1]), "\\d+h$"))

rownames(col_annotation_b) <- colnames(heatmap_df[-1])

# Define colors for CellType
celltype_colors <- c(
  "Macrophage" = "#E41A1C",
  "Monocyte" = "#377EB8", 
  "Neutrophil" = "#4DAF4A",
  "NK" = "#984EA3",
  "pDC" = "#FF7F00",
  "B" = "#FFFF33",
  "T" = "#A65628"
)

# Define colors for Time (gradient from light to dark)
timepoints <- c(
  "00h" = "#ffbe7d",
  "02h" = "#bab0ac",
  "06h" = "#9d7660",
  "12h" = "#d37295",
  "24h" = "#86bcb6",
  "48h" = "#362a39",
  "72h" = "#cd9942"
)

#Combine into annotation colors list
annotation_colors <- list(
  CellType = celltype_colors,
  Time = timepoints
  )

col_annotation_b <- col_annotation_b[, c("Time", "CellType")]

heatmap_matrix <- as.matrix(heatmap_df[-1])


plot3 <- pheatmap(
  heatmap_matrix,
  # silent = TRUE, # If activated will not make any plot in the plot panel!!! but essential for image processing later on 
  annotation_col = col_annotation_b,
  annotation_colors = annotation_colors,
  cluster_rows = T,
  cluster_cols = F,
  color = colorRampPalette(c("white", "darkblue"))(100),
  labels_col = "",
  show_rownames = F,
  #scale = "rows" #this will scale the legend between the positive and negative bounds e.g 6 to -6,
  main = "Heatmap Across Cell Types and Time"
  )
print(plot3)
```


### Task 4. Reproduce panel 2d: Pathway enrichment heatmap
**Goal: Compare pathway-level responses across timepoints.**

Requirements:

    Read sheet d_1
    Use pathway names as rownames
    No clustering
    Diverging color scale centered at zero

Conceptual check:

    Why no clustering here?
    Why a diverging palette?
    
    Response: there is not clustering here because the idea is to check the expression of each pathway at the different timepoint. If clustering is done, it masks the possibility of seeing the biochemistry action response involved in each pathway.
    Resposne: a divergent pallete is helps to distinctly and quickly identify the pathways that are highly upregulated and those that are otherwise.  

```{r Pathway Enrichment Heatmap, message=FALSE, warning=FALSE}
#Read the sheet to use for the task
Pathway_df <- read_excel("hb_stage_2.xlsx", sheet = "d_1")

#Create matrix for the heatmap
pathway_matrix <- as.matrix(Pathway_df[,-1])

#Assign rownames to the dataframe
rownames(pathway_matrix) <- Pathway_df$pathway

plot4 <- pheatmap(
  pathway_matrix,
  silent = TRUE,
  cluster_rows = F,
  cluster_cols = F,
  color = colorRampPalette(c("red", "white", "blue"))(100),
  main = "Pathway Enrichment Heatmap"
  )
print(plot4)
```


### Task 5. Reproduce panel 2e: Bubble plot of kinetic regimes
**Goal: Show count-weighted kinetic behavior.**

Requirements:

    Read sheet e
    Scatter plot:
        x = half_life
        y = alpha
        color = stage
        size = count
    Two legends:
        stage (color)
        count (size)

```{r Bobble Plot of Kinetic Regimes, message=FALSE, warning=FALSE}
#Read the sheet required for the task
bubble_plot_df <- read_excel("hb_stage_2.xlsx", sheet = "e")

plot5 <- ggplot(bubble_plot_df, aes(x = half_life, y = alpha))+
  geom_point(aes(color = stage, size = count), alpha = 0.8)+
  coord_cartesian(
    xlim = c(0,50), ylim = c(0,2)
    )+
  labs(
    x = "Half Life",
    y = "Alpha Life",
    col = "",
    size = "",
    title = "Bobble Plot of Kinetic Regimes: Alpha Life VS Half Life"
  )+
  theme_minimal()+
  theme(
    plot.margin = margin(2,20,2,20),
    plot.title = element_text(hjust = 0.5, margin = margin(t = 12, b = 12)),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    legend.position = c(1,1),
    legend.justification = c(1,1)
  )
print(plot5)
```


### Task 6. Reproduce panel 2f: Stacked proportions
**Goal: Compare B vs Plasma cell proportions over time.**

Requirements:

    Read sheet f
    Subset to s00h and s72h
    Stacked barplot of proportions
    Fixed y-axis (0–0.3)

Conceptual check:

    Why stacked instead of side-by-side?
    
    Response: This is useful to e the relative proportion of each variable in the specific timepoint compared to side by side which might show a difference in magnitude of the variables being considered.

```{r Stacked Barplot, message=FALSE, warning=FALSE}
stacked_bar_df <- read_excel("hb_stage_2.xlsx", sheet = "f")

stacked_bar_df$cell_type <- factor(
  stacked_bar_df$cell_type,
  levels = c("Plasma", "B")
)

stacked_bar_df <- stacked_bar_df %>%
  filter(stage %in% c("s00h", "s72h"))

plot6 <- ggplot(stacked_bar_df,
    aes(x = stage, y = proportion, fill = cell_type)
  )+
  geom_bar(stat = "identity", position = "stack", color = "black")+
  scale_fill_manual(values = c("B" = "#d8a0b3", "Plasma" = "#4f6a9a"))+
  scale_y_continuous(
       breaks = seq(0,0.3, by = 0.05),
       limits = c(0,0.3),
       expand = c(0,0)
     )+
  labs(
    y = "",
    x = "",
    fill = "",
    title = "Stacked Barplot Proportions of Plasma and B cells"
  )+
  theme_minimal()+
  theme(
    plot.margin = margin(2,20,2,20),
    plot.title = element_text(hjust = 0.5, margin = margin(t = 12)),
    panel.grid = element_blank(),
    axis.text = element_text(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
  )
print(plot6)
```



### Task 7. Reproduce panel 2g: Directed cell–cell interaction network
**Goal: Visualize directional communication changes.**

Requirements:

    Read sheet g
    Convert to adjacency matrix
    Build directed graph
    Remove zero-weight edges
    Edge arrow size proportional to weight
    Force-directed layout

Conceptual check:

    Why directed?
    What does edge weight encode biologically?
    
    Response: Directed because this shows the connection and particularly the direction of relationships in the network
    
    Response: Broadly, edge weight encodes the magnitude of interaction between two variables, in our case between the cells. It encodes how strongly or otherwise that two variables estbalish a relationship.

```{r Directed cell–cell Interaction Network, message=FALSE, warning=FALSE}
#Import the sheet required for the task
cell_interaction_df <- read_excel("hb_stage_2.xlsx", sheet = "g")

#Based on the principle of this plot, the matrix must be square i.e nrow() must be same as ncol(). Thus, process the dataframe to meet this requirement then convert the dataframe to matrix
#Convert the tibble to a dataframe
cell_interaction_df <- as.data.frame(cell_interaction_df)

#Assign rownames to the dataframe
rownames(cell_interaction_df) <- cell_interaction_df$...1

#Remove the first column to make it a square data
cell_interaction_df <- cell_interaction_df[-1]

#Now convert to a matrix which is the required data type that the igraph library uses to make the plot
cell_interaction_matrix <- as.matrix(cell_interaction_df)

#Build directed graph and remove zero-weight edges
cell_interaction_graph <- graph_from_adjacency_matrix(
  cell_interaction_matrix,
  mode = "directed", #this parameter ensures direction in the matrix
  weighted = TRUE #this parameter remove zero-weighted edges
)

#Check that weight have been imported
E(cell_interaction_graph)$weight

#Apply force-directed layout using weights
set.seed(123) #Necessary to have reproducible plot otherwise, while the data remains intact, the plot chnages its presentation with every excution of the code.
forced_layout <- layout_with_fr(cell_interaction_graph, weights = E(cell_interaction_graph)$weight)

#Now make the plot with arrow size proportional to weight
png("igraph_plot.png", width = 2000, height = 2000, res = 300)
plot(cell_interaction_graph,
     layout = forced_layout,
     edge.width = E(cell_interaction_graph)$weight,
     edge.arrow.size = 0.5,
     edge.arrow.width = E(cell_interaction_graph)$weight,
     vertex.size = 20,
     vertex.color = "#d8a0b3",
     vertex.label.color = "black",
     vertex.label.cex = 1.5,
     main = "Directed cell–cell Interaction Network"
     )
dev.off()

library(png)
library(grid)
igraph_png <- readPNG("igraph_plot.png")   # load PNG
grid.newpage()
grid.raster(igraph_png)                    # create empty plot
```


### Task 8. Final assembly (mandatory)

    Arrange all panels into a single figure
    Ensure:
        Consistent color usage
        Readable labels
        No clipping
    Export as publication-quality PDF or PNG that can be shared on LinkedIn
    
```{r Final assembly, message=FALSE, warning=FALSE}
# Arrange all panel into a single figure
library(png)
library(grid)
library(gridExtra)
# save each image individually
png("heatmap_plot3.png", width=2000, height=2000, res=300)
grid::grid.draw(plot3$gtable)
dev.off()

png("heatmap_plot4.png", width=2000, height=2000, res=300)
grid::grid.draw(plot4$gtable)
dev.off()

ggsave("ggplot_plot1.png", plot=plot1, width=6, height=6, dpi=300)

ggsave("ggplot_plot2.png", plot=plot2, width=6, height=6, dpi=300)

ggsave("ggplot_plot5.png", plot=plot5, width=6, height=6, dpi=300)

ggsave("ggplot_plot6.png", plot=plot6, width=6, height=6, dpi=300)

# Import the saved images
igraph_img <- rasterGrob(readPNG("igraph_plot.png"))

gg_img1 <- rasterGrob(readPNG("ggplot_plot1.png"))
gg_img2 <- rasterGrob(readPNG("ggplot_plot2.png"))
gg_img5 <- rasterGrob(readPNG("ggplot_plot5.png"))
gg_img6 <- rasterGrob(readPNG("ggplot_plot6.png"))
heatmap_img3 <- rasterGrob(readPNG("heatmap_plot3.png"))
heatmap_img4 <- rasterGrob(readPNG("heatmap_plot4.png"))

# Combine the images
png("combine_figure.png", width = 4000, height = 3000, res = 300)
combined_grob <- arrangeGrob(
  gg_img1,
  gg_img2,
  heatmap_img4,
  gg_img5,
  heatmap_img3,
  gg_img6,
  igraph_img,
  ncol = 4,
  heights = c(1,1)
)

#Display combined image to screen
grid.newpage()
grid.draw(combined_grob) #necessary for drawing and writing the image to file
dev.off()

#Display combined image to screen
combined_img <- readPNG("combine_figure.png")
grid.newpage()
grid.raster(combined_img)
```

Thank you for taking time to follow up to this point. Let's meet in the next lesson: ***Shinny Web Applications in R***